---
title: "Getting Started with ASCENT"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with ASCENT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

ASCENT (Accelerated Single-Cell ENhancer Target gene linking) is a
high-performance reimplementation of the
[SCENT](https://github.com/immunogenomics/SCENT) algorithm. It tests for
statistical association between chromatin accessibility (ATAC-seq peaks) and
gene expression (RNA) across single cells using a GLM with adaptive bootstrap
p-values.

This vignette walks through the same workflow as the original SCENT
`SCENT_interactive` vignette, showing how ASCENT serves as a drop-in
replacement with identical results and significantly faster runtime.

## Load Package

```{r setup}
library(ASCENT)
library(Matrix)
```

## Step 1: Load Inputs

ASCENT requires four inputs:

- **RNA count matrix** (`dgCMatrix`): genes x cells
- **ATAC count matrix** (`dgCMatrix`): peaks x cells
- **Cell metadata** (`data.frame`): must include a `cell` column, covariates,
  and a cell-type column
- **Peak-gene pairs** (`data.frame`): two columns — `gene` and `peak`

```{r load-inputs}
# --- Option A: Load from individual RDS files ---
atac <- readRDS("path/to/atac.rds")       # dgCMatrix, peaks x cells
mrna <- readRDS("path/to/rna.rds")        # dgCMatrix, genes x cells
meta <- readRDS("path/to/meta.rds")       # data.frame with cell metadata
gene_peak <- read.table("path/to/gene_peak_pairs.txt")
colnames(gene_peak) <- c("gene", "peak")

# --- Option B: Extract from a Seurat multiome object ---
# library(Seurat)
# mrna <- GetAssayData(sobj, assay = "RNA",  layer = "counts")
# atac <- GetAssayData(sobj, assay = "ATAC", layer = "counts")
# meta <- data.frame(
#   cell      = colnames(mrna),
#   log_nUMI  = log(sobj$nCount_RNA),
#   log_nATAC = log(sobj$nCount_ATAC),
#   celltype  = sobj$celltype
# )
```

## Step 2: Create an ASCENT Object

The ASCENT S4 object holds all data needed for the analysis. This mirrors
SCENT's `CreateSCENTObj()`.

```{r create-object}
obj <- CreateASCENTObj(
  rna        = mrna,
  atac       = atac,
  meta.data  = meta,
  peak.info  = gene_peak,
  covariates = c("log_nUMI", "log_nATAC"),
  celltypes  = "celltype"
)

# Inspect the object
str(obj, max.level = 2)
head(obj@peak.info)
```

## Step 3: (Optional) Select a Subset of Pairs

For an interactive test run, you can subset to a small number of pairs:

```{r subset-pairs}
obj@peak.info <- obj@peak.info[1:10, ]
head(obj@peak.info)
```

## Step 4: Run the ASCENT Algorithm

`ASCENT_algorithm()` is a drop-in replacement for `SCENT_algorithm()`. The key
addition is the `method` parameter, which selects the computational backend:

- `"rcpp"` (default): Full C++/OpenMP engine. Use this for production runs.
- `"fastglm"`: R-based optimisation using the `fastglm` library. Useful for
  validation.
- `"glm"`: Original SCENT implementation using base `glm()` / `MASS::glm.nb()`.
  Useful as ground-truth reference.

### Poisson regression (default, binarised ATAC)

```{r run-poisson}
result <- ASCENT_algorithm(
  obj,
  celltype = "CD14 Mono",
  ncores   = 8L,
  regr     = "poisson",
  bin      = TRUE,
  method   = "rcpp"
)
```

### Negative binomial regression

```{r run-negbin}
result_nb <- ASCENT_algorithm(
  obj,
  celltype = "CD14 Mono",
  ncores   = 8L,
  regr     = "negbin",
  bin      = TRUE,
  method   = "rcpp"
)
```

### All regression / binarisation combinations

```{r run-all-options}
# Poisson, binarised (default)
res1 <- ASCENT_algorithm(obj, "CD14 Mono", 8L)

# Poisson, non-binarised
res2 <- ASCENT_algorithm(obj, "CD14 Mono", 8L, regr = "poisson", bin = FALSE)

# Negative binomial, binarised
res3 <- ASCENT_algorithm(obj, "CD14 Mono", 8L, regr = "negbin", bin = TRUE)

# Negative binomial, non-binarised
res4 <- ASCENT_algorithm(obj, "CD14 Mono", 8L, regr = "negbin", bin = FALSE)
```

## Step 5: View Results

Results are stored in the `@ASCENT.result` slot (analogous to SCENT's
`@SCENT.result`):

```{r view-results}
res <- result@ASCENT.result

# Sort by bootstrap p-value
res[order(res$boot_basic_p), ]
```

The output columns are:

| Column | Description |
|--------|-------------|
| `gene` | Gene name |
| `peak` | Peak name |
| `beta` | ATAC coefficient from the GLM |
| `se`   | Standard error of the ATAC coefficient |
| `z`    | z-statistic (beta / se) |
| `p`    | Wald p-value (two-sided) |
| `boot_basic_p` | Empirical p-value from adaptive bootstrap |

## Step 6: (Optional) Save Results

```{r save-results}
write.table(result@ASCENT.result,
            file = "ASCENT_result.txt",
            row.names = FALSE, col.names = TRUE, quote = FALSE)
```

## Scaling to Genome-wide Analyses

The original SCENT required a separate parallelisation workflow for genome-wide
analyses: splitting pairs into ~1,000 batches and submitting each as a separate
cluster job (see SCENT's `SCENT_parallelize` vignette). This was necessary
because SCENT parallelises *within* a single pair (bootstrap replicates via
`mclapply` fork), so processing pairs remained sequential.

ASCENT eliminates this complexity. The `rcpp` engine parallelises *across*
pairs via OpenMP — simply increase `ncores` and all pairs are processed in
parallel within a single R session:

```{r genome-wide}
# Generate all cis peak-gene pairs (requires bedtools)
obj <- CreatePeakToGeneList(
  obj,
  genebed = "path/to/GeneBody_500kb_margin.bed",
  nbatch  = 10
)

# Run on all pairs — just increase ncores
obj@peak.info <- do.call(rbind, obj@peak.info.list)
result <- ASCENT_algorithm(obj, "CD14 Mono", ncores = 16L, method = "rcpp")
```

No batch scripts, no job arrays, no result collection step.

For very large analyses (>100k pairs), you can still split into batches
using `obj@peak.info.list` and submit separate SLURM jobs, but far fewer
batches are needed.

## Comparison with SCENT

ASCENT is designed as a drop-in replacement. The table below maps SCENT
functions to their ASCENT equivalents:

| SCENT | ASCENT | Notes |
|-------|--------|-------|
| `CreateSCENTObj()` | `CreateASCENTObj()` | Same arguments |
| `SCENT_algorithm(obj, celltype, ncores)` | `ASCENT_algorithm(obj, celltype, ncores, method = "rcpp")` | Added `method` parameter |
| `CreatePeakToGeneList()` | `CreatePeakToGeneList()` | Identical |
| `obj@SCENT.result` | `obj@ASCENT.result` | Slot name changed |

| | SCENT | ASCENT |
|---|---|---|
| **Parallelism** | Within pair (bootstrap replicates via fork) | Across pairs (OpenMP threads) |
| **Genome-wide workflow** | Split into ~1,000 batches + job arrays | Single call with `ncores > 1` |
| **Memory scaling** | ~N x base per N cores (fork duplication) | Constant regardless of core count |

The `method = "glm"` backend produces numerically identical results to the
original SCENT. The `method = "rcpp"` backend produces equivalent results with faster runtime.
